---
title: "PHP 2550: Project 1"
author: "Peirong Hao"
format: pdf
include-in-header: 
  text: |
    \usepackage{fvextra}
    \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor: visual
execute:
  echo: false
  warning: false
  error: false
  eval: true
  output: true
---

# Introduction

Endurance exercise performance, particularly in long-distance events like marathons, is influenced by a complex interplay of physiological, environmental, and demographic factors. Previous research demonstrates that heat stress slows pacing during aerobic exercise, as it increases core temperature and cardiovascular strain, leading to fatigue and a reduction in overall performance capacity \[1\]. The impact of environmental heat stress is quantified through metrics like the Wet-Bulb Globe Temperature (WBGT), a weighted average of dry bulb, wet bulb, and globe temperature. \[2\] highlights a negative relationship between WBGT and marathon outcomes, with performance declining progressively as WBGT levels rise. These findings underscore the critical role of environmental conditions in determining marathon success.

Age significantly influences performance under heat stress. Older adults face heightened challenges in thermoregulation due to diminished sweat production, reduced cardiovascular efficiency, and lower overall heat dissipation capabilities \[3\]. These limitations impair their ability to maintain performance in warmer conditions, particularly in prolonged endurance activities like marathons. Understanding how age interacts with other environmental variables could provide valuable insight into older athletes' vulnerabilities and needs.

Sex differences add another layer of complexity to endurance performance. While women generally have lower aerobic fitness and maximal oxygen uptake (VO2 max) than men \[4\], they demonstrate distinct advantages, including more consistent pacing strategies and reduced fatigue during endurance running. Even though these characteristics can enhance women's ability to sustain performance over long distances, these benefits are counterbalanced by lower oxygen-carrying capacity and higher body fat percentages, limiting overall performance potential \[5\]. Exploring these differences can inform tailored strategies for improving endurance outcomes across sexes.

This project, conducted in collaboration with Dr. Brett Romano Ely and Dr. Matthew Ely from the Department of Health Sciences at Providence College, aimed to build on these findings by investigating the relationships between environmental factors, age, and sex in marathon performance. Using data from five major marathons collected over the years, this study examined how temperature, humidity, solar radiation, and wind influence marathon performance in men and women. The overarching goal was to provide a more comprehensive understanding of the determinants of marathon success, offering insights to optimize training and environmental adaptation across diverse athlete populations.

# Exploratory Data Analysis (EDA)

```{r}
#| label: setup
#| include: false

set.seed(123456)
library(tidyverse)
library(knitr)
library(tidyr)      
library(dplyr)       
library(kableExtra)
library(readr)
library(visdat)
library(gtsummary)
library(naniar)
#library(patchwork)
library(gridExtra)
library(corrplot)
#library(olsrr)# model selection for linear model
library(splines)
library(lme4)
library(lmerTest)  #p-values for lmer models
library(ggplot2)
library(gridExtra)

#read in data files
df.aqi<-read.csv("aqi_values.csv")
df.course.record<-read.csv("course_record.csv")
df.marathon.dates<-read.csv("marathon_dates.csv")
df.project1.main<-read.csv("project1.csv")
```

```{r}
df.aqi<-df.aqi %>%
  mutate(race = case_when(
          marathon == "Boston" ~ "Boston",
          marathon == "Chicago" ~ "Chicago",
          marathon == "Grandmas" ~ "Grandmas",
          marathon == "NYC" ~ "NYC",
          marathon == "Twin Cities" ~ "TC",
          .default = NA)) %>% 
  select(-c(marathon))  

df.course.record<-df.course.record %>%
  mutate(race = case_when(
          Race == "B" ~ "Boston",
          Race == "C" ~ "Chicago",
          Race == "D" ~ "Grandmas",
          Race == "NY" ~ "NYC",
          Race == "TC" ~ "TC",
          .default = NA)) %>% 
  select(-c(Race))  

df.marathon.dates<-df.marathon.dates %>%
  mutate(race = case_when(
          marathon == "Boston" ~ "Boston",
          marathon == "Chicago" ~ "Chicago",
          marathon == "Grandmas" ~ "Grandmas",
          marathon == "NYC" ~ "NYC",
          marathon == "Twin Cities" ~ "TC",
          .default = NA)) %>% 
  select(-c(marathon))  

df.project1.main<-df.project1.main %>%
  mutate(race = case_when(
          Race..0.Boston..1.Chicago..2.NYC..3.TC..4.D. == 0 ~ "Boston",
          Race..0.Boston..1.Chicago..2.NYC..3.TC..4.D. == 1 ~ "Chicago",
          Race..0.Boston..1.Chicago..2.NYC..3.TC..4.D. == 4 ~ "Grandmas",
          Race..0.Boston..1.Chicago..2.NYC..3.TC..4.D. == 2 ~ "NYC",
          Race..0.Boston..1.Chicago..2.NYC..3.TC..4.D. == 3 ~ "TC",
          .default = NA)) %>% 
  select(-c(Race..0.Boston..1.Chicago..2.NYC..3.TC..4.D.))  
```

```{r}
#| eval: false
# Visualize missing data pattern for all dfs

#aqi column has 16% missingness
df.aqi %>% abbreviate_vars() %>% vis_miss() + theme(text = element_text(size = 7)) + ggtitle("Missing Data")

#no missingness
df.course.record %>% abbreviate_vars() %>% vis_miss() + theme(text = element_text(size = 7)) + ggtitle("Missing Data")

#no missingness
df.marathon.dates %>% abbreviate_vars() %>% vis_miss() + theme(text = element_text(size = 7)) + ggtitle("Missing Data")

#last 8 columns have missingness
df.project1.main %>% abbreviate_vars() %>% vis_miss() + theme(text = element_text(size = 7)) + ggtitle("Missing Data")
```

```{r}
#start to join datasets
df.all <- left_join(df.project1.main, df.course.record, by=c("Year","race"))%>%
  arrange(Year, race) 

#rename sex, age column
df.all<-df.all %>%
  mutate(sex = case_when(
          Sex..0.F..1.M. == 0 ~ "F",
          Sex..0.F..1.M. == 1 ~ "M",
          .default = NA))  %>% 
  select(-c(Sex..0.F..1.M.))  %>%
  rename(age = Age..yr.)

#actual time = CR + %CR * CR -> marathon performance
#CR, unique value for each race, each year
#convert CR from character to time

#https://www.runnersworld.com/races-places/a20794726/age-and-weight-groups-com/
#create age divisions
df.all<-df.all %>%
  mutate(CR = as.numeric(hms(CR)),
    actual.time = CR * (1+ X.CR/100),
    age.division = case_when(
          age <= 14 ~ "14 and under",
          age >= 15 & age <= 19 ~ "15-19",
          age >= 20 & age <= 29 ~ "20-29",
          age >= 30 & age <= 39 ~ "30-39",
          age >= 40 & age <= 49 ~ "40-49",
          age >= 50 & age <= 59 ~ "50-59",
          age >= 60 & age <= 69 ~ "60-69",
          age >= 70 & age <= 79 ~ "70-79",
          age >= 80 ~ "80 and over",
          .default = NA),
        age.division =  factor(age.division, levels = c("14 and under",
        "15-19","20-29","30-39","40-49","50-59","60-69","70-79","80 and over")),
        Flag = factor(Flag, levels = c("White", "Green", "Yellow", "Red", "Black")))

#aqi dataset, parameter_code column: 44201: Ozone, 88101 (primary measure)/88502: PM2.5, multiple rows correspond to one observation
#aqi: air quality index, calculated from mean (integer)
#arithmetic_mean: actual measurement, average sample across that time period, use this instead of aqi; units_of_measure
#do not need "sample_duration"
#WBGT, Flag = 0.7 * Tw + 0.2 * Tg + 0.1 * Td

#merge the other two dfs
df.all <- left_join(df.all, df.marathon.dates, by=c('Year'='year',"race"))%>%
  arrange(Year, race) 

df.aqi$date_local<-as.Date(df.aqi$date_local)
df.all$date<-as.Date(df.all$date)
date.unique.list<-unique(df.all$date)

#pivot wider df.aqi, take mean of arithmetic mean for each region
df.aqi.wider <- df.aqi %>% 
  filter(date_local %in% date.unique.list) %>% 
  #select(date_local, parameter_code, arithmetic_mean, race, units_of_measure) %>% 
  group_by(date_local, race, parameter_code) %>% 
  summarise(mean=mean(arithmetic_mean),.groups = 'drop') %>% 
  pivot_wider(names_from = parameter_code, values_from = mean)

df.all <- left_join(df.all, df.aqi.wider, by=c('date'='date_local',"race"))%>%
  arrange(Year, race) 

df.all <- df.all %>%
  rename(Ozone = `44201`, PM2.5 = `88101`, Sex=sex)
```

```{r}
#| tbl-cap: Summary table of weather parameters and participants' characteristics for all races

df.all %>%
  dplyr::select(-c(Year, X.CR, date, `88502`, age, X.rh, Ozone, PM2.5))%>%
  mutate(actual.time = actual.time/3600, CR=CR/3600)%>%
  tbl_summary(by = race, statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"),
      digits = all_continuous() ~ 2,
      label = c(actual.time ~ "Completion Time (hr)", 
                Wind ~ "Wind speed (km/hr)",
                age.division ~ "Age", 
                Td..C ~ "Dry bulb temperature in Celcius",
                Tw..C ~ "Wet bulb temperature in Celcius",
                Tg..C ~ "Black globe temperature in Celcius",
                SR.W.m2 ~ "Solar radiation in Watts per meter squared",
                DP ~ "Dew Point in Celcius",
                WBGT ~ "Wet Bulb Globe Temperature (WBGT)",
                CR ~ "Course Record(hr)"),
      missing="no"
      #missing_text = "(Missing)",
    ) %>% 
  add_p()%>%
  bold_labels() %>% 
  as_kable_extra(format = "latex", booktabs = TRUE, linesep = "") %>%
  kable_styling(latex_options = "scale_down")
```

Completion times and weather variables varied significantly across races, as reflected by p-values below 0.001 for most characteristics. For example, Grandma's Marathon had the highest average dry bulb temperature at 18.86$°C$, while Boston and NYC reported lower averages of 11.64$°C$ and 11.73$°C$, respectively. Solar radiation also showed notable differences, with Grandma's Marathon recording the highest levels at 676.81 $W/m²$ and NYC the lowest at 401.15 $W/m²$. Wind speed followed a similar trend, with Boston experiencing the highest average at 11.99 $km/hr$, compared to Chicago, which had the lowest at 8.21 $km/hr$. These environmental variations appeared to influence completion times. Runners in Boston achieved the fastest average completion time of 3.3 hours, while participants in New York City had the slowest average time of 3.7 hours.

There were four datasets, including environmental conditions and participants' performances from five major marathons (Boston, Chicago, New York, Twin Cities, and Grandma's Marathons) spanning 1993 to 2016. In total, the data contained 11,564 runners who participated in 98 marathons over the years. As shown in Table 1, weather variables and course records were statistically significantly different across races.

```{r}
#| eval: false
#| fig-cap: Trend plot for completion time
summary_long <- df.all %>%
  group_by(age, Sex) %>%
  summarize(
    Avg = mean(actual.time, na.rm = TRUE)/3600,
    Min = min(actual.time, na.rm = TRUE)/3600,
    Max = max(actual.time, na.rm = TRUE)/3600,
    .groups = "drop") %>%
  pivot_longer(
    cols = c(Avg, Min, Max),
    names_to = "performance_metric",
    values_to = "performance"
  )

ggplot(summary_long, aes(x = age, y = performance, color = Sex, group = interaction(Sex, performance_metric))) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ performance_metric, scales = "free_y", ncol = 1) +
  labs(
    x = "Age Division",
    y = "Course Record (Hours)",
    color = "Sex"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    legend.position = "top",
    panel.grid.major = element_line(color = "gray90")
  )
#On average, as age increases, completion time increase. There was no significant difference between sexes in the overall trend of completion time.
```

```{r}
#| tab-cap: Missingness by Race
miss_race<-df.project1.main %>% filter(if_any(everything(), is.na))%>%
  mutate(race = as.factor(race))%>%
  group_by(Year, race)%>%
  summarise(n = n(),.groups = 'drop')%>%
  rename(Race = race, `Number of observations` = n)
kable(miss_race, booktabs = TRUE, digits = 2)
#MCAR
```

Four matches (2011 Chicago, 2011 NYC, 2011 TC, 2012 Grandmas) had missing weather parameters. Since the missingness was unrelated to other data, it was considered as missing completely at random (MCAR).

```{r}
#| tab-cap: Single variable missingness table in AQI dataset
miss_aqi <- miss_var_summary(df.aqi.wider)[c(1:2),]#first 2 variables have missing values

kable(miss_aqi, format = "latex", booktabs = TRUE, digits = 2)%>%
  kable_styling(latex_options = "hold_position")
```

The air quality index (AQI) dataset contained a significant number of missing values. It is important to note that although PM2.5 was coded in 88502, it had more missing values than 88101, the primary PM2.5 measurement. Because Ozone and PM2.5 measurements were recorded at different sublocations and times of the day for each city, I calculated an average of each parameter for each marathon date and site.

```{r}
#| eval: false
#| tbl-cap: Number of people breaking marathon records by sex as age increases
#| fig-height: 3

trend.record.break <- df.all %>%
  filter(X.CR < 0) %>%
  select(Sex, X.CR, age) %>%
  group_by(age, Sex) %>%
  summarise(n = n(),.groups = 'drop') %>%
  arrange(Sex, age) %>%
  mutate(Sex = factor(Sex)) #%>%
  #pivot_wider(names_from = Sex, values_from = n)

# Trend plot - Break record
ggplot(trend.record.break)+
  geom_line(aes(x=age, y = n, color=Sex, group=Sex), linewidth=1.7)+
    scale_x_continuous(breaks = seq(21, 41, by = 2), guide = guide_axis(angle = 45))+
  theme_minimal()+
  ylab("Number of runners")+
  xlab("Age")
```

The first aim was to examine the effects of increasing age on marathon performance in men and women. CR is the course record for each year's marathon, obtained from last year's record. %CR refers to the percent course record. A positive %CR means the percentage more time used to complete the marathon. A negative %CR means the participant breaks the record. I calculated the completion time for each individual using CR \* (1+%CR). All record-breakers were between the ages of 21 and 41. Males broke the record in a younger age range (21 to 34 years old), but females could break the record in the broader age group (23 to 41 years old).

```{r}
#| fig-cap: Spline plot of completion time vs age
df_clean <- df.all %>%
  filter(!is.na(age) & !is.na(actual.time))

df_clean$completion_time_hours <- df_clean$actual.time / 3600

spline_model_F <- lm(completion_time_hours ~ ns(age, df = 2), data = df_clean[df_clean$Sex == "F", ])
spline_model_M <- lm(completion_time_hours ~ ns(age, df = 2), data = df_clean[df_clean$Sex == "M", ])

age_range <- seq(min(df_clean$age), max(df_clean$age), length.out = 200)

pred_F <- predict(spline_model_F, newdata = data.frame(age = age_range))
pred_M <- predict(spline_model_M, newdata = data.frame(age = age_range))

spline_data <- data.frame(
  age = rep(age_range, 2),
  completion_time = c(pred_F, pred_M),
  sex = rep(c("Female", "Male"), each = length(age_range))
)

ggplot(data = df_clean, aes(x = age, y = completion_time_hours, color = Sex)) +
  geom_point(alpha = 0.2, size = 0.7) +  
  geom_line(data = spline_data, aes(x = age, y = completion_time, color = sex), size = 1.2) +
  labs(
    x = "Age (years)",
    y = "Completion Time (hours)",
    color = "Gender"
  ) +
  scale_color_manual(values = c("Female" = "red", "Male" = "blue")) +
  theme_minimal()

```

Figure 1 reveals a U-shaped relationship between age and marathon completion time. Men generally completed marathons faster than women. People aged between 20 and 40 ran much faster than the other age groups. Regarding the completion time of each match, some people in the elder age group took longer than 9 hours to complete, with the longest time being 11 hours and 54 minutes. These outliers could be due to some data collection error since Boston, Twin Cities, and Chicago marathons must be completed within 6 hours, 6 hours and 15 minutes, 6 and a half hours, respectively. However, Grandma's and NYC marathons allow participants to continue finishing on the sidewalk after 7 hours and 8 hours limit, respectively, without services, including cross-street protection, medical assistance, and aid stations. Thus, even though those outliers weren't official finishing times, they could be valid self-reported completion times.

```{r}
#| eval: false
#| fig-cap: Distribution of completion time for each age group
#| output: true
#| fig-height: 3.5

ggplot(df.all, aes(x=age.division, y=actual.time, fill=Sex)) +
  geom_violin() +
  scale_y_continuous(breaks = c(3600*c(2,3,4,5,6,7,8,9)),
                     labels = c("2HR","3HR","4HR","5HR","6HR","7HR","8HR","9HR"))+
  theme_minimal()+
  labs(x = "Age Group", y = "Completion Time")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

```{r}
#| eval: false
tbl.time<-t(summary(as.period(df.all$actual.time)))
kable(tbl.time, booktabs = TRUE, digits=2)
```

```{r}
#| eval: false

trend.age.impact.summ<-df.all %>%
  select(Sex, actual.time, age) %>%
  group_by(age, Sex) %>%
  summarise(n = n(),.groups = 'drop') %>% 
  # arrange(Sex, age) %>% 
  mutate(Sex = factor(Sex))

trend.age.impact.summ
```

```{r}
#| eval: false

#age: 14-91

ggplot(trend.age.impact.summ)+ 
  geom_line(aes(x=age, y = n, color=Sex, group=Sex), linewidth=1.7)+
    scale_x_continuous(breaks = seq(14, 92, by = 2), guide = guide_axis(angle = 45))+
  ylab("Number of participants")+
  xlab("Age")

df.all %>%
  ggplot(aes(x = age, colour = Sex)) +
  facet_wrap(vars(race), ncol = 3) +
  geom_point(stat = "count",size=1) +
  geom_line(stat = "count") +
  labs(x = "Age", y = "Number of participants")
```

The second aim was to explore the impact of environmental conditions on marathon performance and examine if the effect differs across age and gender. Wet Bulb Globe Temperature (WBGT) is derived from three different sources: Wet bulb temperature (Tw, which accounts for humidity), Black globe temperature (Tg, which accounts for solar radiation), and Dry bub temperature (Td). Flag is determined by WBGT and the chance of heat illness. The Flag variable has five categories arranged from best to worst condition: White, Green, Yellow, Red, and Black. We did not have any data falling in the Black condition, probably because Marathons were canceled in extreme weather.

```{r}
#| fig-cap: Impact of temperature on completion time
#| fig-height: 2.5

df.all %>% ggplot(aes(x=WBGT, y=actual.time/3600, color=age.division)) +
  geom_smooth()+
  theme_minimal()+
  labs(x = "Wet Bulb Globe Temperature (WBGT)", y = "Completion Time in hours",
       color = "Age Group")+
  theme(
    legend.text = element_text(size = 5),
    legend.title = element_text(size = 6)
    )+
  facet_grid(.~Sex)
```

As shown in Figure 2, heat stress, measured by WBGT, significantly impacts marathon performance, with older runners exhibiting more fluctuations in completion times. In contrast, younger and middle-aged runners were less affected, showing relatively stable completion times, though with a slight increasing trend as WBGT rises. While men generally outperform women, both genders exhibit similar patterns of performance.

```{r}
#| fig-cap: Impact of wind on completion time
#| fig-height: 3
df.all %>%
  mutate(race = factor(race, levels=c("Boston", "Grandmas", "TC", "Chicago", "NYC"))) %>%
ggplot(aes(x=Wind, y=actual.time/3600, color=Sex)) +
  geom_smooth()+
  # scale_y_continuous(breaks = c(3600*c(3,4)),
  #                    labels = c("3HR","4HR"))+
  theme_minimal()+
  labs(x = "Wind speed in km/hr", y = "Completion Time in hours")#+
  #facet_grid(.~race)
```

Another important weather parameter is wind speed, measured in $km/h$. In Figure 3, there is a slight difference between the effects of wind speed on completion time for men and women. Women were less affected by changes in wind speed, as indicated by their relatively flat trend line.

```{r}
#| fig-cap: Impact of air pollutants' condition on completion time
#| fig-height: 4

p.ozone<-df.all %>% ggplot(aes(x=Ozone, y=actual.time/3600, color=age.division)) +
  geom_smooth()+
  # scale_y_continuous(breaks = c(3600*c(3,4)),
  #                    labels = c("3HR","4HR"))+
  theme_minimal()+
  labs(x = "Ozone (parts per million)", y = "Completion Time in hours")+
  facet_grid(.~Sex)+
  theme(legend.position="none")

p.pm2.5<-df.all %>% ggplot(aes(x=PM2.5, y=actual.time/3600, color=age.division)) +
  geom_smooth()+
  # scale_y_continuous(breaks = c(3600*c(3,4)),
  #                    labels = c("3HR","4HR"))+
  theme_minimal()+
  labs(x = "PM2.5 (micrograms/cubic meter)", y = "Completion Time in hours",
       color = "Age Group")+
  theme(
    legend.text = element_text(size = 5),
    legend.title = element_text(size = 6)
    )+
  facet_grid(.~Sex)

grid.arrange(p.ozone, p.pm2.5)
```

[Ozone Solutions](https://ozonesolutions.com/ozone-safety/#:~:text=The%20NIOSH%20recommended%20exposure%20limit,dangerous%20to%20life%20or%20health.) suggested the safe limit for ozone is 0.1 ppm, and our Ozone data were well below that threshold. The relationships between Completion Time and Ozone held constant for most age groups but varied slightly for the youngest and eldest groups in Figure 6. According to [Indoor Air Hygiene Institute](https://www.indoorairhygiene.org/pm2-5-explained/#:~:text=Most%20studies%20indicate%20PM2.,breathing%20issues%20such%20as%20asthma.), PM2.5 at or under 12 micrograms/cubic meter is considered healthy, but it is deemed unhealthful when it reaches or surpasses 35 micrograms/cubic meter. Because the recorded PM2.5 data were below 35 micrograms/cubic, there was not much health concern about it. Concerning variation in completion time with PM2.5, there was little change for the majority but slight variation for the youngest and eldest groups. These relations held for both sex groups.

```{r}
#| fig-cap: Correlation plot
#| fig-height: 3

#largest impact analysis: correlation
df.cor.subset<-df.all %>%
  dplyr::select(actual.time, WBGT, Wind, Ozone, PM2.5)

colnames(df.cor.subset) <- c("Completion Time", "WBGT", "Wind", "Ozone", "PM2.5")

corrplot.mixed(cor(df.cor.subset, use = "complete.obs"))
```

The final aim was to identify the weather parameters most affecting marathon performance. No weather parameter showed a strong correlation with completion time, nor were there particularly strong correlations among the weather parameters themselves. Due to the sex differences found earlier, I built two hierarchical random intercept (differ by race) models for females and males. The outcome of interest was `completion time`, and I aimed to find its linear relationship with weather variables, age, and age squared.

```{r}
df_hier <- df.all %>%
  filter(complete.cases(WBGT, Wind, PM2.5, Ozone, Sex, age, actual.time, race)) %>%
  mutate(actual.time=actual.time/3600)   

df_M <- df_hier %>% filter(Sex == "M") %>%
  select(-c(Sex))

df_F <- df_hier %>% filter(Sex == "F") %>%
  select(-c(Sex))

#random intercept models by sex
model_F <- lmer(actual.time ~ poly(age, 2) + WBGT + Wind + PM2.5 + Ozone + 
                          (1|race), data=df_F)

model_M <- lmer(actual.time ~ poly(age, 2) + WBGT + Wind + PM2.5 + Ozone + 
                          (1|race), data=df_M)

df_F$preds <- predict(model_F)
df_M$preds <- predict(model_M)
```

```{r}
kable(summary(model_F)$coefficients, booktabs = TRUE, digits = 3, caption="Fixed Effects - Female")
kable(ranef(model_F)$race, booktabs = TRUE, digits = 3, caption="Random Effects - Female")#Variance: 0.02078
```

```{r}
kable(summary(model_M)$coefficients, booktabs = TRUE, digits = 3, caption="Fixed Effects - Male")
kable(ranef(model_M)$race, booktabs = TRUE, digits = 3, caption="Random Effects - Male")#Variance: 0.02414
```

```{r}
#| fig-cap: Outcome Trend by Weather Parameters
#| fig-height: 9

p.wbgt.f <- ggplot(df_F, aes(x = WBGT, y = preds, color = race)) +
    geom_smooth(method = "lm", fullrange = TRUE, size = 0.3) +
    geom_jitter(aes(x = WBGT, y = actual.time, group = race, color = race),
                alpha = 0.1) +
    labs(x = "WBGT", y = "Completion Time (hr)") +
    ggtitle("Female") +
    scale_colour_discrete('race', name="Race")


p.wbgt.m <- ggplot(df_M, aes(x = WBGT, y = preds, color = race)) +
    geom_smooth(method = "lm", fullrange = TRUE, size = 0.3) +
    geom_jitter(aes(x = WBGT, y = actual.time, group = race, color = race),
                alpha = 0.1) +
    labs(x = "WBGT", y = "Completion Time (hr)") +
    ggtitle("Male") +
    scale_colour_discrete('race', name="Race")


p.wind.f <- ggplot(df_F, aes(x = Wind, y = preds, color = race)) +
    geom_smooth(method = "lm", fullrange = TRUE, size = 0.3) +
    geom_jitter(aes(x = Wind , y = actual.time, group = race, color = race),
                alpha = 0.1) +
    labs(x = "Wind ", y = "Completion Time (hr)") +
    ggtitle("Female") +
    scale_colour_discrete('race', name="Race")


p.wind.m <- ggplot(df_M, aes(x = Wind, y = preds, color = race)) +
    geom_smooth(method = "lm", fullrange = TRUE, size = 0.3) +
    geom_jitter(aes(x = Wind, y = actual.time, group = race, color = race),
                alpha = 0.1) +
    labs(x = "Wind", y = "Completion Time (hr)") +
    ggtitle("Male") +
    scale_colour_discrete('race', name="Race")


p.pm2.5.f <- ggplot(df_F, aes(x = PM2.5, y = preds, color = race)) +
    geom_smooth(method = "lm", fullrange = TRUE, size = 0.3) +
    geom_jitter(aes(x = PM2.5, y = actual.time, group = race, color = race),
                alpha = 0.1) +
    labs(x = "PM2.5", y = "Completion Time (hr)") +
    ggtitle("Female") +
    scale_colour_discrete('race', name="Race")


p.pm2.5.m <- ggplot(df_M, aes(x = PM2.5, y = preds, color = race)) +
    geom_smooth(method = "lm", fullrange = TRUE, size = 0.3) +
    geom_jitter(aes(x = PM2.5, y = actual.time, group = race, color = race),
                alpha = 0.1) +
    labs(x = "PM2.5", y = "Completion Time (hr)") +
    ggtitle("Male") +
    scale_colour_discrete('race', name="Race")


p.ozone.f <- ggplot(df_F, aes(x = Ozone , y = preds, color = race)) +
    geom_smooth(method = "lm", fullrange = TRUE, size = 0.3) +
    geom_jitter(aes(x = Ozone, y = actual.time, group = race, color = race),
                alpha = 0.1) +
    labs(x = "Ozone", y = "Completion Time (hr)") +
    ggtitle("Female") +
    scale_colour_discrete('race', name="Race")  + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


p.ozone.m <- ggplot(df_M, aes(x = Ozone , y = preds, color = race)) +
    geom_smooth(method = "lm", fullrange = TRUE, size = 0.3) +
    geom_jitter(aes(x = Ozone , y = actual.time, group = race, color = race),
                alpha = 0.1) +
    labs(x = "Ozone", y = "Completion Time (hr)") +
    ggtitle("Male") +
    scale_colour_discrete('race', name="Race")  + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

grid.arrange(p.wbgt.f, p.wbgt.m, 
             p.wind.f, p.wind.m,
             p.pm2.5.f, p.pm2.5.m,
             p.ozone.f, p.ozone.m, nrow = 4)
```

In terms of fixed effects, the intercept and coefficients for `age`, `age squared`, `WBGT`, `wind`, and `PM2.5` are similar in both the female and male models. Although the coefficient for `ozone` differs a little between the two models, the p-values indicate that the estimated coefficients for `ozone` are not statistically significant. The interpretation of the significant weather parameters for the female model is as follows: for each additional unit increase in WBGT, the completion time is expected to increase by 0.006 hours, holding all other variables constant. For each additional unit increase in wind speed, the completion time is expected to increase by 0.008 hours, holding all other variables constant. For each additional unit increase in PM2.5, the completion time is expected to increase by 0.009 hours, holding all other variables constant. The interpretation for the male model is similar. The random effects for both female and male models are also quite similar, with the variance of the intercept being 0.02078 and 0.02414, respectively. The coefficient values are relatively small, and their magnitudes are similar across female and male groups, as illustrated in Figure 6.

# Conclusion

People aged between 20 and 40 run much faster than the other age groups and could break course records. Age and weather parameters have a positive linear relationship with completion time: older runners finish the marathon later; severe weather, such as strong winds and higher temperatures, also delays completion. Additionally, we discovered that, on average, men run faster than women. Ozone is not s statistically significant parameter in the random intercept models. This could result from the observed ozone level falling between 0 and 0.054 ppm, which is thought to be good air quality ([Kaiterra](https://learn.kaiterra.com/en/resources/ozone-what-levels-are-safe)). Four marathons had incomplete weather data. If we conduct imputation for missing values, we might get more robust results. It is also worth examining more marathons in different regions to investigate the impact of weather on performance further.

# References

\[1\] Ely, B. R., Cheuvront, S. N., Kenefick, R. W., & Sawka, M. N. (2010). Aerobic performance is degraded, despite modest hyperthermia, in hot environments. Med Sci Sports Exerc, 42(1), 135-41.

\[2\] Ely, M. R., Cheuvront, S. N., Roberts, W. O., & Montain, S. J. (2007). Impact of weather on marathon-running performance. Medicine and science in sports and exercise, 39(3), 487-493.

\[3\] Kenney, W. L., & Munce, T. A. (2003). Invited review: aging and human temperature regulation. Journal of applied physiology, 95(6), 2598-2603.

\[4\] Yanovich, R., Ketko, I., & Charkoudian, N. (2020). Sex differences in human thermoregulation: relevance for 2020 and beyond. Physiology, 35(3), 177-184.

\[5\] Besson, T., Macchi, R., Rossi, J., Morio, C. Y., Kunimasa, Y., Nicol, C., ... & Millet, G. Y. (2022). Sex differences in endurance running. Sports medicine, 52(6), 1235-1257.

\newpage

# Code Appendix

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
